Efficient simulation of individual-based population models

### 2024-12-10 article.qmd
computo revision before publication

---
title: "Efficient simulation of individual-based population models"
subtitle: "The `R` package `IBMPopSim`"
author:
  - name: Daphné Giorgi 
    corresponding: true
    email: daphne.giorgi@sorbonne-universite.fr
    url: https://perso.lpsm.paris/~giorgi/
    affiliations:
      - name: Sorbonne Université, CNRS
        department: Laboratoire de Probabilités, Statistique et Modélisation
        url: https://www.lpsm.paris/
  - name: Sarah Kaakai
    corresponding: true
    email: sarah.kaakai@univ-lemans.fr
    url: https://perso.univ-lemans.fr/~skaakai/
    affiliations:
      - name: Le Mans Université
        department: Laboratoire Manceau de Mathématiques
        url: https://lmm.univ-lemans.fr/fr/
      - name: Ecole Polytechnique, CNRS
        department: Centre de Mathématiques Appliquées
        url: https://cmap.ip-paris.fr/ 
  - name: Vincent Lemaire 
    corresponding: true
    email: vincent.lemaire@sorbonne-universite.fr
    url: https://perso.lpsm.paris/~vlemaire/
    orcid: 0000-0002-0433-7722
    affiliations:
      - name: Sorbonne Université, CNRS
        department: Laboratoire de Probabilités, Statistique et Modélisation
        url: https://www.lpsm.paris/
date: last-modified
date-modified: last-modified
description: |
    This document provides a full description of the Stochastic Individual-Based Models (IBMs) that can be implemented in the IBMPopSim package. A unified mathematical and simulation framework is given, with a detailed description of the simulation algorithm. Examples of applications for the package are also provided, showing the performance and flexibility of IBMPopSim. 
abstract: >+
    The `R` Package `IBMPopSim` facilitates the simulation of the random evolution of heterogeneous populations using stochastic Individual-Based Models (IBMs). The package enables users to simulate population evolution, in which individuals are characterized by their age and some characteristics, and the population is modified by different types of events, including births/arrivals, death/exit events, or changes of characteristics. The frequency at which an event can occur to an individual can depend on their age and characteristics, but also on the characteristics of other individuals (interactions). Such models have a wide range of applications in fields including  actuarial science, biology, ecology or epidemiology. `IBMPopSim` overcomes the limitations of time-consuming IBMs simulations by implementing new efficient algorithms  based on thinning methods, which are compiled using the `Rcpp` package while providing a user-friendly interface.
keywords: [Individual-based models, stochastic simulation, population dynamics, Poisson measures, thinning method, actuarial science, insurance portfolio simulation]
citation:
  type: article-journal
  container-title: "Computo"
  doi: "xxxx"
  url: https://computo.sfds.asso.fr/template-computo-quarto
  publisher: "Société Française de Statistique"
  issn: "2824-7795"
bibliography: references.bib
github-user: VincentLemaire
repo: "ibmpopsim_computo"
draft: true # set to false once the build is running
published: false # will be set to true once accepted
format:
    computo-html: default
    computo-pdf: default
crossref:
    cnj-title: Assumption
    cnj-prefix: Assumption 
execute:
    eval: true
    echo: true
    warning: false
---

@@

Since the seminal paper of @FouMel04, it has been shown that a stochastic IBM dynamics can be defined rigorously as the unique solution of an SDE driven by Poisson measures, under reasonable non explosion conditions. This representation has actually been proved only in particular case, for various models. We provide here a general mathematical framework.  
Some recalls on Poisson random measures are presented in the Appendix @sec-recall-poisson, and for more details on these representations of particular examples, we refer to the abundant literature on the subject (see [@bremaud1981point, @Cin11] and the references therein).

@@

![Realization of a Marked Poisson measure $\bar Q$ on $[0,T]$ with mean measure $\bar \mu(\mathrm{d} t, \mathrm{d} \theta) = \mathrm{d} t \mathbf{1}_{[0, \bar \Lambda]}(\theta) \mathrm{d} \theta$ (red crosses), and realization of the restriction $\bar Q^\Delta$ where $\Delta = \{ (t,\theta)\in [0,T]\times[0, \bar \Lambda], \theta \leq \Lambda(t) \}$ (blue circles). The projection of $\bar Q^\Delta$ on first component is an inhomogeneous Poisson process on $[0,T]$ of intensity $(\Lambda(t))$ and jump times $(T_k)_{k \ge 1}$.](img/plot_thinning.tex){#fig-thinning}

@@

```pseudocode
#| label: alg-randomized
#| html-indent-size: "1.2em"
#| html-comment-delimiter: "//"
#| html-line-number: true
#| html-line-number-punc: ":"
#| html-no-end: false
#| pdf-placement: "H"
#| pdf-line-number: true

@@

# Appendix {.appendix}

## Recall on Poisson random measures {#sec-recall-poisson .appendix}

@@

![Example of Marked Poisson measure on $[0,T]$ with $m(\mathrm{d} t) = L \mathrm{d} t$ (jump times occur at Poisson arrival times of rate $L$) and with $\nu(\mathrm{d} y) = \frac{1}{L} \mathbf{1}_{[0, L]}(y) \mathrm{d} y$ (marks are drawn uniformly on $[0,L]$).
The mean measure is then $\mu(\mathrm{d} t, \mathrm{d} y) = \mathrm{d} t {1}_{[0,L]}(y) \mathrm{d} y$.](img/plot_poisson.tex){#fig-poisson}

## Pathwise representation of IBMs {#sec-pathwise-representation}

@@

## Proof of @thm-interaction {#sec-prf-thm-interaction}

@@

## Acknowledgements {.appendix}

>>>>

---
title: "Efficient simulation of individual-based population models"
subtitle: "The `R` package `IBMPopSim`"
author:
  - name: Daphné Giorgi 
    corresponding: true
    email: daphne.giorgi@sorbonne-universite.fr
    url: https://perso.lpsm.paris/~giorgi/
    affiliations:
      - name: Sorbonne Université, CNRS
        department: Laboratoire de Probabilités, Statistique et Modélisation
        url: https://www.lpsm.paris/
  - name: Sarah Kaakai
    corresponding: true
    email: kaakai@math.univ-paris13.fr
    url: https://www.math.univ-paris13.fr/~kaakai/
    affiliations:
      - name: Université Sorbonne Paris Nord, CNRS
        department: Laboratoire Analyse, Géométrie et Applications
        url: https://www.math.univ-paris13.fr/
      - name: Ecole Polytechnique, CNRS
        department: Centre de Mathématiques Appliquées
        url: https://cmap.ip-paris.fr/ 
  - name: Vincent Lemaire 
    corresponding: true
    email: vincent.lemaire@sorbonne-universite.fr
    url: https://perso.lpsm.paris/~vlemaire/
    orcid: 0000-0002-0433-7722
    affiliations:
      - name: Sorbonne Université, CNRS
        department: Laboratoire de Probabilités, Statistique et Modélisation
        url: https://www.lpsm.paris/
date: last-modified
date-modified: last-modified
description: |
    This document provides a full description of the Stochastic Individual-Based Models (IBMs) that can be implemented in the IBMPopSim package. A unified mathematical and simulation framework is given, with a detailed description of the simulation algorithm. Examples of applications for the package are also provided, showing the performance and flexibility of IBMPopSim. 
abstract: >+
    The `R` Package `IBMPopSim` facilitates the simulation of the random evolution of heterogeneous populations using stochastic Individual-Based Models (IBMs). The package enables users to simulate population evolution, in which individuals are characterized by their age and some characteristics, and the population is modified by different types of events, including births/arrivals, death/exit events, or changes of characteristics. The frequency at which an event can occur to an individual can depend on their age and characteristics, but also on the characteristics of other individuals (interactions). Such models have a wide range of applications in fields including  actuarial science, biology, ecology or epidemiology. `IBMPopSim` overcomes the limitations of time-consuming IBMs simulations by implementing new efficient algorithms  based on thinning methods, which are compiled using the `Rcpp` package while providing a user-friendly interface.
keywords: [Individual-based models, stochastic simulation, population dynamics, Poisson measures, thinning method, actuarial science, insurance portfolio simulation]
citation:
  type: article-journal
  container-title: "Computo"
  doi: "xxxx"
  url: https://computo.sfds.asso.fr/template-computo-quarto
  publisher: "Société Française de Statistique"
  issn: "2824-7795"
bibliography: references.bib
github-user: VincentLemaire
repo: "ibmpopsim_computo"
draft: true # set to false once the build is running
published: false # will be set to true once accepted
format:
    computo-html: default
    computo-pdf: default
crossref:
    cnj-title: Assumption
    cnj-prefix: Assumption 
execute:
    eval: true
    echo: true
    warning: false
---

@@

Since the seminal paper of @FouMel04, it has been shown that a stochastic IBM dynamics can be defined rigorously as the unique solution of an SDE driven by Poisson measures, under reasonable non explosion conditions. This representation has actually been proved only in particular case, for various models. We provide here a general mathematical framework.  
Some recalls on Poisson random measures are presented in the Appendix @sec-recall-poisson, and for more details on these representations of particular examples, we refer to the abundant literature on the subject (see [@bremaud1981point; @Cin11] and the references therein).

@@

![Realization of a Marked Poisson measure $\bar Q$ on $[0,T]$ with mean measure $\bar \mu(\mathrm{d} t, \mathrm{d} \theta) = \mathrm{d} t \mathbf{1}_{[0, \bar \Lambda]}(\theta) \mathrm{d} \theta$ (red crosses), and realization of the restriction $\bar Q^\Delta$ where $\Delta = \{ (t,\theta)\in [0,T]\times[0, \bar \Lambda], \theta \leq \Lambda(t) \}$ (blue circles). The projection of $\bar Q^\Delta$ on first component is an inhomogeneous Poisson process on $[0,T]$ of intensity $(\Lambda(t))$ and jump times $(T_k)_{k \ge 1}$.](img/plot_thinning.png){#fig-thinning}

@@

```pseudocode
#| label: alg-randomized
#| html-indent-size: "1.2em"
#| html-comment-delimiter: "//"
#| html-line-number: true
#| html-line-number-punc: ":"
#| html-no-end: false
#| pdf-placement: "htb!"
#| pdf-line-number: true

@@

# Appendix

## Recall on Poisson random measures {#sec-recall-poisson }

@@

![Example of Marked Poisson measure on $[0,T]$ with $m(\mathrm{d} t) = L \mathrm{d} t$ (jump times occur at Poisson arrival times of rate $L$) and with $\nu(\mathrm{d} y) = \frac{1}{L} \mathbf{1}_{[0, L]}(y) \mathrm{d} y$ (marks are drawn uniformly on $[0,L]$).
The mean measure is then $\mu(\mathrm{d} t, \mathrm{d} y) = \mathrm{d} t {1}_{[0,L]}(y) \mathrm{d} y$.](img/plot_poisson.png){#fig-poisson}

## Pathwise representation of IBMs {#sec-pathwise-representation  }

@@

## Proof of @thm-interaction {#sec-prf-thm-interaction  }

@@

## Acknowledgements
